Left-Linear Rewriting in Adhesive Categories

### 2024-07-06 concur/final.tex
minimal e funding

\funding{The research has been partially supported by the Italian MUR under the PNRR PRIN 2022P2022HXNSC ``RAP''
and the PRIN 20228KXFN2 ``STENDHAL'' and by the University of Pisa under the PRA 2022\_99 ``FM4HD''.}

@@

So far, most of the theoretical results focused on what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens already
in the context of string diagrams mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07}, or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expressions, the idea
here -- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

@@

In this paper we perform a systematic exploration of these phenomena at the
% Here we perform this exploration at the
level of $\mathcal{M}$-adhesive categories,
which encompass a large family of DPO-based formalisms adopted by researchers and 
practitioners alike.

@@

On the one hand, they capture most categories of
interest for rewriting. Indeed, we identify quite general classes of
left-linear rewriting systems on graph-like structures that are
well-switching.
% and we show that rewriting systems on graphs with
% equivalences are so.
As a sanity check, we prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, by recovering, 
even if sometimes in
weakened form, most of the results that hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development has to face a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely
\begin{enumerate}
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent;
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result does not depend on the order in which switchings are applied, but
  only on the associated permutation.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

@@

%It can be seen that 
$\mathcal{M}$-adhesivity of $\X$ guarantees the
uniqueness of the result of applying a rule to an object, i.e.~two
direct derivations using the same match are abstraction equivalent
(see Proposition~\ref{prop:unique}).

@@

Still, there are $\mathcal{M}$-adhesive rewriting systems that are not strong enforcing, as shown by the following example.

>>>>

\funding{The research has been partially supported by the EuropeanUnion - NextGenerationEU under the National Recovery and Resilience Plan (NRRP) - Call PRIN 2022 PNRR - Project P2022HXNSC ``Resource Awareness in Programming: Algebra, Rewriting, and Analysis'', by the Italian MUR - Call PRIN 2022 - Project 20228KXFN2 
``Spatio-Temporal Enhancement of Neural nets for Deeply
Hierarchical Automatised Logic'' and by the University of Pisa - Call PRA 2022 - Project 2022\_99 ``Formal Methods for the Healthcare Domain based on Spatial Information''.}

@@

So far, most of the theoretical results focused on %what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens
in the context of string diagrams mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07}, or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
merging nodes corresponding to equal expressions, the idea 
-- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

@@

In this paper we perform a systematic exploration of these phenomena at the
% Here we perform this exploration at the
level of $\mathcal{M}$-adhesive categories,
which encompass a large family of DPO-based formalisms.
% adopted by researchers and  practitioners alike.

@@

On the one hand, they capture most categories of
interest for rewriting. Indeed, we identify general classes of
left-linear rewriting systems on graph-like structures that are
well-switching.
% and we show that rewriting systems on graphs with
% equivalences are so.
As a sanity check, we prove that linear rewriting systems 
%on $\mathcal{M}$-adhesive categories 
are well-switching.

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, 
sometimes in
weakened form, most of the results that hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development faces a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely
\begin{enumerate}
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent;
%\newpage
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result does not depend on the order in which switchings are applied, but
  only on the associated permutation.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

@@

%It can be seen that 
$\mathcal{M}$-adhesivity of $\X$ ensures the
uniqueness of the result of applying a rule to an object: two
direct derivations using the same match are abstraction equivalent
(see Proposition~\ref{prop:unique}).

@@

Still, there are $\mathcal{M}$-adhesive rewriting systems that are not strong enforcing.
%as shown by the following example.
